Visual narratives of the COVID-19 pandemic

### 2022-10-28 JDSSV_manuscript_submitted.Rnw
Tweaks for copyediting

At the same time, we still see a lot of defective graphics disseminated and shared; some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. 
These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others \citep{EQUATORNetworkEnhancing}. 
The use of numbers, their graphic representation and the systematic creation of charts and tables from data in the mass media have been accompanied from the beginning by efforts to improve the communication of quantitative and statistical information. And while efforts continue to be made in this area, historical documents describe many of the same problems we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
In other words, while our \textit{technology} for recording, visually summarising, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when \textit{designing} charts which communicate data accurately and effectively.

@@

Yet another approach to showing the temporal component of the pandemic was to leverage web graphics to show the temporal component through animation; this freed up the $x$ axis which in the default configuration would have been devoted to time and allowed another variable to be displayed. While this approach was common on social media, it was reserved for specific use-cases and more carefully selected variables by professional media outlets. \Cref{fig:bubble-ft} shows an example from the Financial Times which uses the x-axis to show GDP per capita and the y-axis to show vaccination rate, bubble size indicates a country's population. With so many variables and ratios of variables, these charts are complex to read and use, even though they are visually appealing.

@@

There are many different ways that designers can use time series data to provide additional contextual information and facilitate comparisons. \Cref{fig:ft-waves} shows COVID cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant. This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends. In addition, it is clear that the chart is highlighting the Omicron variant from the colour selection: the red used to show the Omicron numbers is a sharp contrast from the green and blue shades used to show the first three waves of the pandemic. Notably, even though red and green in combination are usually to be avoided for the sake of colourblind viewers, the shades of blue/green variants which are chosen are light enough that the red stands out even to someone with colour vision deficiency.

@@

First ideas have been presented at DSSV 2021. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.

>>>>

At the same time, we still see a lot of defective graphics disseminated and shared; some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. 
These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others \citep{EQUATORNetworkEnhancing}. 
The use of numbers, their graphic representation and the systematic creation of charts and tables from data in the mass media have been accompanied from the beginning by efforts to improve the communication of quantitative and statistical information. While efforts continue to be made in this area, historical documents describe many of the same problems we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
In other words, while our \textit{technology} for recording, visually summarising, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when \textit{designing} charts which communicate data accurately and effectively.

@@

Yet another approach to showing the temporal component of the pandemic was to leverage web graphics to show the temporal component through animation; this freed up the $x$ axis which in the default configuration would have been devoted to time and allowed another variable to be displayed. While this approach was common on social media, it was used more selectively, with carefully chosen variables by professional media outlets. \Cref{fig:bubble-ft} shows an example from the Financial Times which uses the x-axis to show GDP per capita and the y-axis to show vaccination rate, bubble size indicates a country's population. With so many variables and ratios of variables, these charts are complex to read and use, even though they are visually appealing.

@@

There are many different ways that designers can use time series data to provide additional contextual information and facilitate comparisons. \Cref{fig:ft-waves} shows COVID cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant. This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends. In addition, it is clear that the chart is highlighting the Omicron variant from the colour selection; the red used to show the Omicron numbers is a sharp contrast from the green and blue shades used to show the first three waves of the pandemic. Notably, even though red and green in combination are usually to be avoided for the sake of colourblind viewers, the shades of blue/green variants which are chosen are light enough that the red stands out even to someone with colour vision deficiency.

@@

These ideas were first presented at DSSV 2020. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.
